# Life and Economic Loss by Severe Weather events

## Synopsis


In this report we try to find the changes in population health loss and economic loss by the severe weather events in the United States between the years 1951 and 2011. Our overall hypothesis is that severe weather events  has made huge impact in people life and countries economy. To investigate this hypothesis, we obtained event data from the NOAA Storm Database which is collected from monitors sited across the U.S. We specifically obtained data for the years from 1951 and 2011. From these data, we found that, on average across the U.S., thousands of people lost  their life and injured and billion of dollars damaged all these years.   In particular two events are very dangerous to life and economy. Also these severe weather events makes disaster by an average of every year.

```{r message=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
```

## Reading and Processing the Raw Data

Read the NOAA Storm Data as a compressed csv file and examine the data.

```{r}
storm_data <- read.csv("repdata_data_StormData.csv.bz2")
dim(storm_data)

head(storm_data[, 1:15])
```

Then collect only the losses data and see the few data
```{r}
storm_data <- storm_data %>% 
          select (BGN_DATE,EVTYPE, FATALITIES, INJURIES, PROPDMG,PROPDMGEXP, CROPDMG, CROPDMGEXP  )

storm_data$BGN_DATE <- as.Date(storm_data$BGN_DATE, '%m/%d/%Y %H:%M:%S')
storm_data$year <- year(storm_data$BGN_DATE)


levels(storm_data$PROPDMGEXP)<- c("1", "1","1","1","1","1",
                                   "1","1","1","1","1","1","1", "1000000000",
                                   "100", "100", "1000", "1000000", "1000000")
levels(storm_data$CROPDMGEXP)<- c("1", "1", "1", "1", "100000000", "1000", "1000", "1000000", "1000000")

storm_data$PROPDMGEXP  <-as.numeric(levels(storm_data$PROPDMGEXP))[storm_data$PROPDMGEXP]
storm_data$CROPDMGEXP<-as.numeric(levels(storm_data$CROPDMGEXP))[storm_data$CROPDMGEXP]

storm_data$PROPDMG<- (storm_data$PROPDMG * storm_data$PROPDMGEXP)
storm_data$CROPDMG<-(storm_data$CROPDMG * storm_data$CROPDMGEXP)

storm_data$TOTALDMG<-(storm_data$CROPDMG + storm_data$PROPDMG)

head(storm_data)


```

Lets aggragate the data by the event types
```{r}
aggrated_data <- storm_data  %>% 
                  select( FATALITIES, INJURIES, PROPDMG, CROPDMG,TOTALDMG, EVTYPE) %>% 
                              group_by(EVTYPE) %>% summarise_all(funs(sum))
```

Examine the data now
```{r}
head(aggrated_data)
summary(aggrated_data)
```

Also aggragate the data by the year
```{r}
yearly_aggragated_data <- storm_data  %>% 
          select( FATALITIES, INJURIES, PROPDMG, CROPDMG, year) %>% 
          group_by(year) %>% summarise_all(funs(sum))
     
```

Examine yearly aggragated the data
```{r}
head(yearly_aggragated_data)
```

## Results

### The total Fatalities and Injuries by the events those are greater than average threat

```{r}
aggrated_data$health <- aggrated_data$FATALITIES + aggrated_data$INJURIES  

par(mar=c(14, 4, 4, 6))
barplot(aggrated_data$health[aggrated_data$health > mean(aggrated_data$health)] , names.arg=aggrated_data$EVTYPE[aggrated_data$health > mean(aggrated_data$health)],
        main="The total Fatalities and Injuries by the severe events",
        xlab="Events", ylab="Health Related Loss ( Fatalities + Injuries)",
        las=2, cex.lab=0.25)
```

### Population Loss by the TORNADO alone by every year

```{r}
tornado_data <- storm_data  %>% group_by(year) %>% filter(EVTYPE == 'TORNADO') %>% gather(TYPE, LOSS, FATALITIES:INJURIES)

     ggplot(tornado_data, aes(x=year, y=LOSS, colour=TYPE))+
                geom_point()+
                facet_wrap(~TYPE)+
                labs(x = "Year", 
                     y = "Number of Losses", 
                     title="The total Fatalities and Injuries by TORNADO")
```

### The Property and Agriculture Demage by the events those are greater than average threat

```{r}
par(mar=c(14, 4, 4, 8))
barplot(aggrated_data$TOTALDMG[aggrated_data$TOTALDMG > mean(aggrated_data$TOTALDMG)] , names.arg=aggrated_data$EVTYPE[aggrated_data$TOTALDMG > mean(aggrated_data$TOTALDMG)],
        main="The total Property and Agriculture demage by Severe Events",
        xlab="Events", ylab="Economic Loss ( In thousands )",
        las=2, cex.lab=0.25)
```

### Which event caused most demage
#### 1. Fatalities
```{r}
     aggrated_data[which.max(aggrated_data$FATALITIES ),]
```
#### 2. Injuries
```{r}
     aggrated_data[which.max(aggrated_data$INJURIES ),]
```

#### 3. Property demage
```{r}
     aggrated_data[which.max(aggrated_data$PROPDMG ),]
```

#### 4. Agriculture demage
```{r}
     aggrated_data[which.max(aggrated_data$CROPDMG ),]
```

### The total losses by all these years
#### 1. Popluation Health ( Fatalities and Injuries )
```{r}
     sum(aggrated_data$health)
```
    
#### 2. Property demage ( in thousands)
```{r}
     sum(aggrated_data$PROPDMG)
```

#### 2. Agriculture demage ( in thousands )
```{r}
     sum(aggrated_data$CROPDMG)
```


### Yearly aggragation of all events
```{r}
     yearly_aggragated_data <- storm_data  %>% select( FATALITIES, INJURIES, PROPDMG, CROPDMG, year) %>% group_by(year) %>% summarise_all(funs(sum))
     head(yearly_aggragated_data)
     
     summary( yearly_aggragated_data)
```


It is clear that the severe weather events are making very significant damage to life and economy, in particular Tornado, Flood and Drought. Good prediction and precaution on such weather events can reduce the losses.
